Live Functional Programming with Typed Holes

### 2018-04-16 appendix-defns.tex
revisions of sec3 through 3.4 and additional proofs and definitions in the appendix

\vspace{5pt}
\begin{lem}[Substitution] \label{thm:substitution}~
\begin{enumerate}[nolistsep]
\item If $\hasType{\hDelta}{\hGamma, x : \htau'}{d}{\htau}$ and $\hasType{\hDelta}{\hGamma}{d'}{\htau'}$ then $\hasType{\hDelta}{\hGamma}{[d'/x]d}{\htau}$.
\item If $\hasType{\hDelta}{\hGamma, x : \htau'}{\sigma}{\hGamma'}$ and $\hasType{\hDelta}{\hGamma}{d'}{\htau'}$ then $\hasType{\hDelta}{\hGamma}{[d'/x]\sigma}{\hGamma'}$.
\end{enumerate}
\end{lem}

@@

@@

\clearpage
\subsection{Hole Filling}\label{sec:hole-filling}
\input{fig-multi-step}
\input{fig-evalctx-instantiation}
\begin{lem}[Filling] ~
  \begin{enumerate}[nolistsep]
  \item If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\dexp}{\tau}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
  \item If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\sigma}{\hGamma''}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\sigma}}{\hGamma''}$.
  \end{enumerate}
\end{lem}
\begin{proof}
In each case, we proceed by rule induction on the first assumption, appealing to the Substitution Lemma as necessary.
\end{proof}

\begin{lem}[Substitution Commutativity]
  If 
  \begin{enumerate}[nolistsep]
  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{x : \htau_2}{\dexp_1}{\tau}$ and 
  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_2}{\htau_2}$ and 
  	\item $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  \end{enumerate}

  then  $\instantiate{d'}{u}{\substitute{d_2}{x}{d_1}} = \substitute{\instantiate{d'}{u}{d_2}}{x}{\instantiate{d'}{u}{d_1}}$.
\end{lem}

@@

  then $\reducesE{}{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{lem}

\begin{lem}[Filling Totality]
Either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx}=\evalctx'$ for some $\evalctx'$.
\end{lem}

@@

	then $\instantiate{d}{u}{d_1} = \instantiate{d}{u}{d_2}$.
\end{lem}

\begin{lem}[Filling Distribution] If 
	$\selectEvalCtx{d_1}{\evalctx}{d_1'}$ and $\instantiate{d}{u}{\evalctx}=\evalctx'$ then $\selectEvalCtx{\instantiate{d}{u}{d_1}}{\evalctx'}{\instantiate{d}{u}{d_1'}}$.
\end{lem}

@@

  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{thm}
\begin{proof}
By rule induction on (3). The reflexive case is immediate. In the inductive case, we proceed by rule induction on the stepping premise. There is one rule, Rule~{Step}. By Filling Totality, either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx} = \evalctx'$. In the former case, by Discarding, we can conclude by MultiStepRefl. In the latter case, by Instruction Commutativity and Filling Distribution we can take a step, and we can conclude via MultiStepSteps by applying Preservation and then the IH.
\end{proof}

>>>>

\vspace{5pt}
\begin{lem}[Substitution] \label{thm:substitution}~
\begin{enumerate}[nolistsep]
\item If $\hasType{\hDelta}{\hGamma, x : \htau'}{d}{\htau}$ and $\hasType{\hDelta}{\hGamma}{d'}{\htau'}$ then $\hasType{\hDelta}{\hGamma}{[d'/x]d}{\htau}$.
\item If $\hasType{\hDelta}{\hGamma, x : \htau'}{\sigma}{\hGamma'}$ and $\hasType{\hDelta}{\hGamma}{d'}{\htau'}$ then $\hasType{\hDelta}{\hGamma}{[d'/x]\sigma}{\hGamma'}$.
\end{enumerate}
\end{lem}
\begin{proof} By rule induction on the first assumption in each case. The conclusion follows from the definition of substitution in each case. \end{proof}

@@

The proofs for all three of these theorems follow by straightforward rule induction.

@@

\subsection{Multiple Steps}
\input{fig-multi-step}

\clearpage
\subsection{Hole Filling}\label{sec:hole-filling}
\begin{lem}[Filling] ~
  \begin{enumerate}[nolistsep]
  \item If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\dexp}{\tau}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\dexp}}{\tau}$.
  \item If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\hGamma}{\sigma}{\hGamma''}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  then $\hasType{\hDelta}{\hGamma}{\instantiate{\dexp'}{u}{\sigma}}{\hGamma''}$.
  \end{enumerate}
\end{lem}
\begin{proof}
In each case, we proceed by rule induction on the first assumption, appealing to the Substitution Lemma as necessary.
\end{proof}

We need the following auxiliary definitions, which lift hole filling to evaluation contexts taking care to consider the special situation where the mark is inside the hole that is being filled, to prove the Commutativity theorem.
\input{fig-evalctx-instantiation}
\begin{lem}[Substitution Commutativity]
  If 
  \begin{enumerate}[nolistsep]
  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{x : \htau_2}{\dexp_1}{\tau}$ and 
  	\item $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_2}{\htau_2}$ and 
  	\item $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$
  \end{enumerate}

  then  $\instantiate{d'}{u}{\substitute{d_2}{x}{d_1}} = \substitute{\instantiate{d'}{u}{d_2}}{x}{\instantiate{d'}{u}{d_1}}$.
\end{lem}
\begin{proof}
We proceed by structural induction on $d_1$ and rule induction on the typing premises, which serve to ensure that the free
variables in $d'$ are accounted for by every closure for $u$.
\end{proof}

@@

  then $\reducesE{}{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{lem}
\begin{proof}
We proceed by cases on the instruction transition assumption (no induction is needed). For Rule \rulename{ITBeta}, we defer to the Substitution Commutativity lemma above. For the remaining cases, the conclusion follows from the definition of hole filling.
\end{proof}

\begin{lem}[Filling Totality]
Either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx}=\evalctx'$ for some $\evalctx'$.
\end{lem}
\begin{proof} We proceed by structural induction on $\evalctx$. Every case is handled by one of the two judgements. \end{proof}

@@

	then $\instantiate{d}{u}{d_1} = \instantiate{d}{u}{d_2}$.
\end{lem}
\begin{proof} We proceed by structural induction on $\evalctx$ and rule induction on all three assumptions. Each case follows from the definition of instruction selection and hole filling. \end{proof}

\begin{lem}[Filling Distribution] If 
	$\selectEvalCtx{d_1}{\evalctx}{d_1'}$ and $\instantiate{d}{u}{\evalctx}=\evalctx'$ then $\selectEvalCtx{\instantiate{d}{u}{d_1}}{\evalctx'}{\instantiate{d}{u}{d_1'}}$.
\end{lem}
\begin{proof} We proceed by rule induction on both assumptions. Each case follows from the definition of instruction selection and hole filling. \end{proof}

@@

  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{thm}
\begin{proof}
By rule induction on assumption (3). The reflexive case is immediate. In the inductive case, we proceed by rule induction on the stepping premise. There is one rule, Rule~\rulename{Step}. By Filling Totality, either $\inhole{u}{\evalctx}$ or $\instantiate{d}{u}{\evalctx} = \evalctx'$. In the former case, by Discarding, we can conclude by \rulename{MultiStepRefl}. In the latter case, by Instruction Commutativity and Filling Distribution we can take a \rulename{Step}, and we can conclude via \rulename{MultiStepSteps} by applying Filling, Preservation and then the induction hypothesis.
\end{proof}
